SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-04-10 2017-Xudong-SMT/ActaInformatica.tex
some minor changes

\abstract{Open pNets are formal models that can express the behaviour of open systems, either
	synchronous, asynchronous, or heterogeneous. They are endowed with a symbolic operational semantics in
	terms of open automata, which allows us to check properties of
	such systems in a compositional manner. 
	We present an algorithm computing these semantics,
	building predicates expressing the synchronisation conditions
	between the events of the pNet sub-systems. Checking such
	predicates requires symbolic reasoning about first order logics and application-specific data. We use the Z3 SMT engine to check
	satisfiability of the predicates.
	%As an industrial oriented use-case, %we use so-called ``architectures''
	%for BIP systems, that have been used in the framework of an ESA project and to specify the control software of a nanosatellite at the EPFL Space Engineering Center. 
%	We also present several use-cases, in particular, encoding a BIP architecture
%	extended with explicit data and computing its
%	open automaton semantics. This automaton may be used to prove
%	behavioural properties such as safety and liveness.
We also propose and implement an optimised
	algorithm that performs part of the pruning on the fly, and show 
	its correctness with respect to the original one.
	We illustrate the approach using two use-cases: the first one is a
	classical process-algebra operator for which we provide several
	encodings, and prove some basic properties. The second one is  industry-oriented and based on the so-called ``BIP architectures'', which have
	been used to specify the
	control software of a nanosatellite at the EPFL Space Engineering
	Center. We use pNets to encode a BIP architecture
	extended with explicit data, compute its semantics and discuss its
	properties, and then show how our algorithms scale up, using a
	composition of two such architectures.}

@@

Parameterised Networks of Synchronised Automata (pNets) were proposed
to give a behavioural specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. They are used in
VerCors~\cite{HKLM:Foclasa14}, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalise
each component of a distributed system and their composition. The model of
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterised labelled transition systems (PLTS), where each
hierarchical level defines the possible synchronisation of the lower levels.
Traditionally, pNets have been used to formalise fully
defined systems or softwares. But we  also want to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet model addresses this problem, using
\emph{holes} as process parameters, representing \emph{unspecified} subsystems.
The open pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} in which many examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons.
The operational semantics of an open pNet is defined as an
open automaton in which open transitions contain logical predicates
expressing the relations between the behaviour of the holes and the
global behaviour of a system. In previous work,
only a sketch of a procedure allowing for computing these semantics was
presented, together with a proof of finiteness of the open automaton, under
reasonable hypotheses on the pNet structure.

@@

\paragraph{Contribution.}
In the current work we contribute in the following aspects.
\begin{itemize}
	\item We present an open automaton generation algorithm. We
	implement a full working prototype, within the 
	VerCors platform. In the process, we improve the
	semantic rules from~\cite{henrio:Forte2016}, and add features in
	the algorithm to deal with the full 
	model, including management of variables and assignments.
	\item We implement the interaction between our algorithm and the Z3
	SMT solver, for checking the satisfiability of the transitions
	generated by the algorithm.
	\item We define a strategy checking satisfiability of open transitions in the course of the residual algorithm, and show that  the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
	\item We show the usage of this approach on various use-cases from different settings, including an
	industry-inspired case-study, namely one architectural pattern
	extracted (and extended) from the BIP specification of a nanosatellite on-board
	software.
\end{itemize}

@@

For other applications, such as the analysis of 
programming languages, there exist dedicated platforms using
external automatic theorem 
provers (ATP) or automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from a user model. 
The prover we use, which also happens to be Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories framework (SMT), and is mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011} or others.

@@

\paragraph{Structure.}
In Section
\ref{section:pnets} we give a description and a formal definition of
the pNet model. 
In Section \ref{section:op-semantics} we briefly recall  the operational semantics
of pNets.
In Section \ref{section:implementation} we present an algorithm to compute this semantics, including the interaction with Z3.
In Section~\ref{section:use-cases} we give a few use-cases, in particular, a BIP
architecture from the nanosatellite case-study. 
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

pNets can encode naturally the notion of input actions as found,
\eg in value-passing CCS 
\cite{Milner89} or of usual point-to-point message passing calculi,
but it also allows 
for more general mechanisms, like gate negotiation in LOTOS~\cite{LotosISO89}, or broadcast
communications.

@@

	\begin{itemize}
		\item $Sorts$ is a set of data sorts.
		% \noteSB{Commented out a footnote here\mdash check if still necessary.}
		%% \footnote{Later we may want to extend this with Sort constructors,
		%%   like Array or Pair, but this is not needed now}
		\item $\mathit{Constrs}$ is a set of \emph{constructor operators}: for each $\mathit{Con} \in \mathit{Constrs}$, $arity(Con)=n \in \mathbb{N}$ is its arity 
		and '$Con : (\mathit{sel}_1,\mathit{sort}_1), \dots, (\mathit{sel}_n,\mathit{sort}_n) \rightarrow \mathit{sort}$' is its signature with the associated selectors.
		For each argument, the pair $(\mathit{sel}_i,\mathit{sort}_i)$ defines an auxiliary
		operator of name $\mathit{sel}_i$ with signature $\mathit{sel}_i : \mathit{sort} \rightarrow \mathit{sort}_i$.
		\item $Ops$ is a set of \emph{auxiliary operators}, with their
		arity and signature, of the form: $Op : \mathit{sort}_1, \dots,  \mathit{sort}_n \rightarrow
		\mathit{sort}$.
		\item $\mathit{Constrs}(\mathit{sortname})$ and $\mathit{Sels}(\mathit{sortname})$ are, respectively, the sets of
		constructors and selectors of the sort $\mathit{sortname}$.
	\end{itemize}
	%  \noteSB{I think this can be removed}\hl{All sorts and operator names must be distinct.}
	Constructors of arity 0 are called \emph{constants}, and denoted by $\mathit{Consts}(\APres)$.
	%  $\mathit{Consts}(\APres) = \{\mathit{Con} \in \mathit{Constrs}\,|\, \mathit{arity}(\mathit{Con})=0\}$.
\end{definition}

Sorts \texttt{Bool} and \texttt{Int} are predefined with standard operators.
Sort \texttt{Action} is similar, with a constructor \texttt{Synchro} denoting
a synchronised action, i.e. an ``internal'' action that cannot be
further synchronised with the environment. It also comes with an
overloaded \texttt{FUN} constructor, used to build actions with
arguments, which will be instantiated to the required sorts for a given
pNet.

@@

\begin{table}[t]\caption{\label{table:BIPalgebra}\leftline{Algebra
			Presentation: predefined Sorts and Operators}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
		\\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
		$\texttt{true},\ \texttt{false}$&
		$\land,\ \lor,\ \neg,\ \implies,\ =,\ \ne$
		\\\specialrule{0em}{1pt}{1pt} 
		%               		Action 			&  $\texttt{Synchro},\ \texttt{FUN\_Action\_Bool}$ &
		Action 			&  $\texttt{Synchro},\ \texttt{FUN}$ &
		\\\specialrule{0em}{1pt}{1pt}
		Int 				&
		${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
		$- \texttt{(unary)},\ +,\ -
		\texttt{(binary)},\ \times,\ \div \text{ \etc}$
		%                \\\specialrule{0em}{1pt}{1pt}
		%                \textsl{for any sort} & & $=,\ \ne$
		\\\hline\specialrule{0em}{1pt}{1pt}
		\multicolumn{3}{l}{\sl Extension for the Enable use-case of Section~\ref{section:running-example}}
		\\\hline\specialrule{0em}{1pt}{1pt}
		Action & \texttt{ACT}, \texttt{delta, acc}  &
		\\\hline
	\end{tabular}

@@

		\item[$\bullet$] $\symb{SV}_k^{k\in K}$ is a set of
		synchronisation vectors ($K\in\I_\P$).  It is required that \linebreak $\forall k\!\in\! K,
		\symb{SV}_k\!=\!\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k\,|\,g_k$, where
		$\alpha'_k\in \actions{\variables}$, $I_k\subseteq I$, $J_k\subseteq J$, and 
		$\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
			J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
		$\alpha'_k$. The Boolean expression $g_k $, with $\vars(g_k)\subseteq \bigcup_{l\in 
			I_k\uplus J_k}{\vars({\alpha_l})}$, is a guard associated to the vector.

@@

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition represents the composition transitions 
of several
% \noteSB{These should all be PLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; a transition occurs if its predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions.
%Strictly speaking, the LTSs at the leaves of the
%open automaton are a restricted form of PLTSs, where labels are
%parametrised actions, but include no guard nor assignments.

@@

\paragraph{Predicates.}
Let
$\langle\overline{\pNet},J,\symb{SV}_k^{k\in K} \rangle$
be a pNet. Consider a synchronisation vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\begin{multline*}
	\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Longleftrightarrow
	%% \begin{array}{l}
	\exists {(a'_i)}^{i\in I},
	{(b'_j)}^{j\in J},v'.\\
	SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
	\land
	\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v'
\end{multline*}

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_1^0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_1^n}$=RefineReachableSubAutomaton($\mathit{OA_1^{n-1}}$)
\Until  $\mathit{OA_1^n}=\mathit{OA_1^{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_1^n}$;

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is naturally not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and checked for
satisfiability (function \texttt{filterUnsatTransitions}). 
Then we compute (lines 12-14) the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote by $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet (obtained in line~3 of the algorithm); %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of $n$-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$.

@@

Function \texttt{filterUnsatTransitions}, in line 11 of
Algorithm\ref{alg1}, and function \\
\texttt{RefineReachableSubAutomaton}
in line 13, make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate, and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) value of
variable the state variable $v$.

@@

The basic algorithm from the previous section potentially explores some
areas of the global state space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state-space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models, as illustrated by the discussion above. By avoiding building
  the OTs coming from some unreachable states, we also avoid adding
  the corresponding assignments to the variables of their target states.
But we \emph{cannot} use the
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. We first need to compute a first over-approximation of
the reachable state space and assignment sets; this corresponds to
the automaton $OA_2^0$ in Algorithm~\ref{alg3}, line 13.
So the smart algorithm proceeds in two steps: during the
first step, the SAT checking is done without using any assignment set
information (line 5).  Only the OTs found SAT from this check will be added
to $L$, and the corresponding target global state added to $U$ if
necessary. This ensures that all explored global states are
effectively reachable by a chain of SAT OTs.

@@

At the end of the residual loop, we do another traversal of the
automaton, this time using the assignment information collected that
far (line 12), and compute the final reachable sub-automaton
iteratively (lines 13-15), in the same way than in algorithm \ref{alg1}. 
Naturally, this step is only useful if there are some assignments.

\medskip
Now we will prove that the smart algorithm computes the same 
open automaton.

@@

\begin{theorem}[Correctness without assignments]
  \label{Th:CorrectnessWithoutAssignments}
For a pNet verifying the finiteness conditions of theorem 1, and
containing no assignment to state variables, then the smart  
algorithm, without a final reachability computation, computes 
the same open automaton as that of Algorithm 1 up to a renaming of
variables in the transition labels.  
  \end{theorem}

\begin{proof}[Theorem \ref{Th:CorrectnessWithoutAssignments}]

>>>>

\abstract{Open pNets are formal models that can express the behaviour of open systems, either
	synchronous, asynchronous, or heterogeneous. They are endowed with a symbolic operational semantics in
	terms of open automata, which allows us to check properties of
	such systems in a compositional manner. 
	We present an algorithm computing these semantics,
	building predicates expressing the synchronisation conditions
	between the events of pNet sub-systems. Checking such
	predicates requires symbolic reasoning about first order logics and application-specific data. We use the Z3 SMT engine to check
	satisfiability of the predicates.
	%As an industrial oriented use-case, %we use so-called ``architectures''
	%for BIP systems, that have been used in the framework of an ESA project and to specify the control software of a nanosatellite at the EPFL Space Engineering Center. 
%	We also present several use-cases, in particular, encoding a BIP architecture
%	extended with explicit data and computing its
%	open automaton semantics. This automaton may be used to prove
%	behavioural properties such as safety and liveness.
We also propose and implement an optimised
	algorithm that performs part of the pruning on the fly, and show 
	its correctness with respect to the original one.
	We illustrate the approach using two use-cases: the first one is a
	classical process-algebra operator for which we provide several
	encodings, and prove some basic properties. The second one is  industry-oriented and based on the so-called ``BIP architectures'', which have
	been used to specify the
	control software of a nanosatellite at the EPFL Space Engineering
	Center. We use pNets to encode a BIP architecture
	extended with explicit data, compute its semantics and discuss its
	properties, and then show how our algorithms scale up, using a
	composition of two such architectures.}

@@

Parameterised networks of synchronised automata (pNets) were proposed
to give a behavioural specification formalism for distributed
systems, synchronous, asynchronous, or heterogeneous. They are used in
VerCors~\cite{HKLM:Foclasa14}, a platform for designing and 
verifying distributed systems, as the intermediate language for various
high-level languages. The high-level languages in VerCors formalise
each component of a distributed system and their composition. The model of
pNets provides the core low-level semantic formalism for VerCors, and
is made of a hierarchical composition of (value-passing) automata,
called parameterised labelled transition systems (PLTS), where each
hierarchical level defines the possible synchronisation of the lower levels.
Traditionally, pNets have been used to formalise fully
defined systems or softwares. But we  also want to define and reason
about incompletely defined systems, like program skeletons, operators,
or open expressions of process calculi.
The open pNet model addresses this problem, using
\emph{holes} as process parameters, representing \emph{unspecified} subsystems.
The open pNet model was developed in a series of
papers~\cite{HMZ:PDP15,henrio:Forte2016} in which many examples have been
introduced showing its ability to encode the operators from some
other algebras or  program skeletons.
The operational semantics of an open pNet is defined as an
open automaton in which open transitions contain logical predicates
expressing the relations between the behaviour of the holes and the
global behaviour of a system. In previous work,
only a sketch of a procedure allowing for computing these semantics was
presented, together with a proof of finiteness of the open automaton, under
reasonable hypothesis on the pNet structure.

@@

\paragraph{Contribution.}
In the current work we contribute in the following aspects.
\begin{itemize}
	\item We present an open automaton generation algorithm. We
	implement a full working prototype, within the 
	VerCors platform. In the process, we improve the
	semantic rules from~\cite{henrio:Forte2016}, and add features in
	the algorithm to deal with the full 
	model, including management of variables and assignments.
	\item We implement the interaction between our algorithm and the Z3
	SMT solver, for checking the satisfiability of the transitions
	generated by the algorithm.
	\item We define a strategy checking satisfiability of open transitions in the course of the residual algorithm, and show the correctness of the optimised algorithm w.r.t. the original one.
	% the reachable part of the open automaton can be generated and it is minimal with respect to the decidability of predicate satisfiability.
	\item We show the usage of this approach on various use-cases from different settings, including an
	industry-inspired case-study, namely one architectural pattern
	extracted (and extended) from the BIP specification of a nanosatellite on-board
	software.
\end{itemize}

@@

For other applications, such as the analysis of 
programming languages, there exist dedicated platforms using
external automatic theorem 
provers (ATPs) or automatic tactics from interactive theorem
provers (ITPs), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from a user model. 
The prover we use, which also happens to be Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories framework (SMT), and is mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011} or others.

@@

\paragraph{Structure.}
In Section
\ref{section:pnets} we give a description and a formal definition of
the pNet model. 
In Section \ref{section:op-semantics} we briefly recall  the operational semantics
of pNets.
In Section \ref{section:implementation} we present an algorithm to compute this semantics, including the interaction with Z3.
In Section~\ref{section:use-cases} we give a few use-cases, in particular, a BIP
architecture from the nanosatellite case-study. 
Finally, we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

pNets can encode naturally the notion of input actions as found,
\eg in value-passing CCS 
\cite{Milner89} or of usual point-to-point message passing calculi,
but they also allow 
for more general mechanisms, like gate negotiation in LOTOS~\cite{LotosISO89}, or broadcast
communications.

@@

	\begin{itemize}
		\item $Sorts$ is a set of data sorts.
		% \noteSB{Commented out a footnote here\mdash check if still necessary.}
		%% \footnote{Later we may want to extend this with Sort constructors,
		%%   like Array or Pair, but this is not needed now}
		\item $\mathit{Constrs}$ is a set of \emph{constructor operators}: for each $\mathit{Con} \in \mathit{Constrs}$, $arity(Con)=n \in \mathbb{N}$ is its arity 
		and $Con : (\mathit{sel}_1,\mathit{sort}_1), \dots, (\mathit{sel}_n,\mathit{sort}_n) \rightarrow \mathit{sort}$ is its signature with the associated selectors.
		For each argument, the pair $(\mathit{sel}_i,\mathit{sort}_i)$ defines an auxiliary
		operator of name $\mathit{sel}_i$ with signature $\mathit{sel}_i : \mathit{sort} \rightarrow \mathit{sort}_i$.
		\item $Ops$ is a set of \emph{auxiliary operators}, with their
		signatures in the form: \\ $Op : \mathit{sort}_1, \dots,  \mathit{sort}_n \rightarrow
		\mathit{sort}$.
		\item $\mathit{Constrs}(\mathit{sortname})$ and $\mathit{Sels}(\mathit{sortname})$ are, respectively, the sets of
		constructors and selectors of the sort $\mathit{sortname}$.
	\end{itemize}
	%  \noteSB{I think this can be removed}\hl{All sorts and operator names must be distinct.}
	Constructors of arity 0 are called \emph{constants}, and denoted by $\mathit{Consts}(\APres)$.
	%  $\mathit{Consts}(\APres) = \{\mathit{Con} \in \mathit{Constrs}\,|\, \mathit{arity}(\mathit{Con})=0\}$.
\end{definition}

In Table~\ref{table:BIPalgebra} we give an algebra presentation.
Sorts \texttt{Bool} and \texttt{Int} are predefined with standard operators.
Sort \texttt{Action} is similar, with a constructor \texttt{Synchro} denoting
a synchronised action, i.e. an ``internal'' action that cannot be
further synchronised with the environment. It also comes with an
overloaded \texttt{FUN} constructor, used to build actions with
arguments, which will be instantiated to the required sorts for a given
pNet.

@@

\begin{table}[t]\caption{\label{table:BIPalgebra}\leftline{Algebra
			presentation: predefined sorts and operators}}
	\begin{tabular}{p{2.8cm}p{2.8cm}p{5cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
		\\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
		$\texttt{true},\ \texttt{false}$&
		$\land,\ \lor,\ \neg,\ \implies,\ =,\ \ne$
		\\\specialrule{0em}{1pt}{1pt} 
		%               		Action 			&  $\texttt{Synchro},\ \texttt{FUN\_Action\_Bool}$ &
		Action 			&  $\texttt{Synchro},\ \texttt{FUN}$ &
		\\\specialrule{0em}{1pt}{1pt}
		Int 				&
		${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
		$- \texttt{(unary)},\ +,\ -
		\texttt{(binary)},\ \times,\ \div \text{ \etc}$
		%                \\\specialrule{0em}{1pt}{1pt}
		%                \textsl{for any sort} & & $=,\ \ne$
		\\\hline\specialrule{0em}{1pt}{1pt}
		\multicolumn{3}{l}{\sl Extension for the Enable use-case of Section~\ref{section:running-example}}
		\\\hline\specialrule{0em}{1pt}{1pt}
		Action & \texttt{ACT}, \texttt{delta, acc}  &
		\\\hline
	\end{tabular}

@@

		\item[$\bullet$] $\symb{SV}_k^{k\in K}$ is a set of
		synchronisation vectors ($K\in\I_\P$).  It is required that each
		$\symb{SV}_k$ is a pair $(\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k,\ g_k)$, where
		$\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
			J_k}{\vars({\alpha_l})}$, $\alpha'_k\in \actions{\variables}$, $I_k\subseteq I$, and $J_k\subseteq J$. The global action of a vector $\symb{SV}_k$ is
		$\alpha'_k$. The Boolean expression $g_k $, with $\vars(g_k)\subseteq \bigcup_{l\in 
			I_k\uplus J_k}{\vars({\alpha_l})}$, is a guard associated to the vector.

@@

The semantics of an open pNet will be defined as an open automaton, which
is an automaton where each transition represents the composition transitions 
of several
% \noteSB{These should all be PLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; a transition occurs if its predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions.
%Strictly speaking, the LTSs at the leaves of the
%open automaton are a restricted form of PLTSs, where labels are
%parametrised actions, but include no guard nor assignments.

@@

\paragraph{Predicates.}
Let
$\langle\overline{\pNet},J,\symb{SV}_k^{k\in K} \rangle$
be a pNet. Consider a synchronisation vector $SV_k$, for $k\in K$. We build a
predicate $\MkPred$ relating
the actions of the involved sub-pNets and the resulting actions. This predicate verifies:
\begin{multline*}
	\MkPred(SV_k, a_i^{i\in I}, b_j^{j\in J}, v)\Longleftrightarrow
	%% \begin{array}{l}
	\exists {(a'_i)}^{i\in I},
	{(b'_j)}^{j\in J},v'.\\
	SV_k={(a'_i)}^{i\in I}, {(b'_j)}^{j\in J}\rightarrow v'
	\land
	\forall i\in I.\, a_i=a'_i\land \forall j \in J.\, b_j=b'_j \land v=v' . 
\end{multline*}

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=\textit{filterUnsatTransitions}($\mathit{L,Assignments}$);
\State $\mathit{OA_1^0}$=\textit{makeReachableSubAutomaton}($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_1^n}$=\textit{RefineReachableSubAutomaton}($\mathit{OA_1^{n-1}}$)
\Until  $\mathit{OA_1^n}=\mathit{OA_1^{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_1^n}$;

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is naturally not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and then checked for
satisfiability (function \textit{filterUnsatTransitions}). 
Then we compute (lines 12-14) the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote by $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet (obtained in line~3 of the algorithm); %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of $n$-tuples, is the
cartesian product:  
$$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n). $$

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{RefineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) value of
 the state variable $v$.

@@

The basic algorithm introduced above potentially explores some
areas of the global state space that will not be reachable after
checking the OTs' predicates for satisfiability. Incorporating the
satisfiability check inside the residual algorithm will definitely
save some of these useless computation. More precisely, it will still
create, and submit to Z3, the OTs at the ``border'' of the state space, but
not the transitions issued from non-reachable global states. We use
this idea to create our ``smart'' algorithm.

But there is more to this. Due to the symbolic nature of the
  transitions, reachability is not as easy to define as in standard
  models, as illustrated by the discussion above. By avoiding building
  the OTs coming from some unreachable states, we also avoid adding
  the corresponding assignments to the variables of their target states.
But we \emph{cannot} use the
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. We first need to compute an over-approximation of
the reachable state space and assignment sets; this corresponds to
the automaton $OA_2^0$ in Algorithm~\ref{alg3}, line 13.
So the smart algorithm proceeds in two steps: during the
first step, the SAT checking is done without using any assignment set
information (line 5).  Only the OTs found SAT from this check will be added
to $L$, and the corresponding target global state added to $U$ if
necessary. This ensures that all explored global states are
effectively reachable by a chain of SAT OTs.

@@

At the end of the residual loop, we do another traversal of the
automaton, this time using the assignment information collected that
far (line 12), and compute the final reachable sub-automaton
iteratively (lines 13-15), in the same way as in Algorithm \ref{alg1}. 
Naturally, this step is only useful if there are some assignments.

\medskip
Now we will prove that the smart algorithm computes essentially the same 
open automata as those of the original algorithm.

@@

\begin{theorem}[Correctness without assignments]
  \label{Th:CorrectnessWithoutAssignments}
For a pNet satisfying the finiteness conditions of Theorem~\ref{Th:finiteness} and
containing no assignment to state variables, the smart  
algorithm, without a final reachability computation, computes 
the same open automaton as that of Algorithm~\ref{alg1} up to a renaming of
variables in the transition labels.  
  \end{theorem}

\begin{proof} %[Theorem \ref{Th:CorrectnessWithoutAssignments}]

 Let $pn$ be a pNet. The set of states 
 managed by the two algorithms both are subsets of the cartesian
 product of the PLTS states, as defined in page \pageref{def-states}. In a
 similar way, the set of state variables attached to each state are
 also equal. Comparing  
 transitions requires a little more care, because within the
 $MakeTransitions$ function, fresh variables are created everytime a
 synchronisation vector is used (this occurs within the $MkPred$ and
 $matching$ functions). However, given a global state, the call of
 $MakeTransitions$ will always return exactly the same set of OTs, up to a
renaming of these fresh variables. When this is not ambiguous, we will
speak of equality (of transitions, and of open automata) without the phrase ``up to a renaming of fresh variables''.
